---
output:
  html_document: default
  output: github_document
  pdf_document: default
author: "Joerg Heintz"
date: "3/9/2017"
title: "Reproducible Research Project Week 2"
---

###Setting global options
```{r setoptions, echo = TRUE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/joergheintz/Documents/08_MPHPHI/11_Coursera/Coursera_ReproduciableResearch/week2")
```

###Read "activity.csv"
```{r}
mydata<-read.csv("activity.csv", header = TRUE)
```

###Total number of steps, histogram, mean, and median
The data set contains NA, and can introducte bias.

#####Total number of steps taken per day
```{r,}
totSteps_I<-sum(mydata$steps, na.rm = TRUE)
totSteps_II<-sum(tapply(mydata$steps, mydata$date, sum, na.rm=TRUE))
```
The total number of steps taken by this persons within the month period were `r totSteps_I` steps. 

#####Histogram
The histogram illustrates the frequency of steps per day [NA considered na.rm=FALSE].
```{r,  fig.width=5, fig.height=3}
hist(tapply(mydata$steps, mydata$date, sum, na.rm=FALSE),
     main = "Frequency of steps taken per day",
     xlab = "Steps per day",
     ylab = "Frequency",
     breaks = 20,
     border = "blue",
     col="lightblue", 
     xlim=c(0,25000),
     ylim=c(0, 12),
     las=2)
```


#####Mean, Median of the total number of steps taken per day
The following table is based on the total number of steps = `r totSteps_I` . 
```{r,  results='asis'}
library(xtable)
StepsByDay<-tapply(mydata$steps , mydata$date, sum, na.rm=TRUE)
stepsMean<-round(mean(StepsByDay), 0)
stepsMedian<-median(StepsByDay)
totSteps_I<-sum(mydata$steps, na.rm = TRUE)
df<-data.frame()
df<-rbind(stepsMean, stepsMedian,totSteps_I)
df<-cbind(df, c(" mean per day ", " median per day ", " total all days "))
colnames(df)<-c(" Values ", " Description ")
xt <- xtable(df)
print(xt, type = 'html', include.rownames = FALSE)
```


#Average daily activity pattern
```{r, echo=TRUE, fig.width=5, fig.height=3}
library(data.table)
library(ggplot2)
StepsByInterval<-data.frame(tapply(mydata$steps, mydata$interval, mean, na.rm = TRUE))
StepsByInterval <-cbind(StepsByInterval, data.frame(rownames(StepsByInterval)))
colnames(StepsByInterval)<-c("steps", "interval")
x<-as.numeric(as.character(StepsByInterval$interval))
y<-as.numeric(as.character(StepsByInterval$steps))
IntMax<-as.numeric(as.character(StepsByInterval[StepsByInterval$steps == max(StepsByInterval$steps), 2]))
plot(x,y,type="l", col = "red", xlab = "Interval", ylab = "Average Steps")
abline(v = IntMax, col = "blue", lty = "dotted")
#text((800), min(y), "number", pos=2)
axis(1, at=IntMax,labels=IntMax, col = "blue")
```

###Interval with max average steps
```{r, echo=TRUE}
IntMax<-StepsByInterval[StepsByInterval$steps == max(StepsByInterval$steps), 2]
```
The `r IntMax` interval is the 5-minute interval, that contains on across all the days the maximum average number of steps. 


##Imputing missing values
#### Total number of rows with NAs
```{r, echo = TRUE}
myNAdata<-sum(is.na(mydata))
```

`r myNAdata` is the total number of NA in the data set.

#### Strategy for filling in missing values (NA's)
The missing values will filled with mean values of the corresponsing intervals.

#### New Data Set that is equal to the original dataset but with the missing data filled in
```{r, echo = FALSE}
# get data
mydata<-read.csv("activity.csv", header = TRUE)
# calc mean per interval
IntMean<-round(data.frame(tapply(mydata$steps, mydata$interval, mean, na.rm=TRUE)), 0)
IntMean<-cbind(data.frame(rownames(IntMean)), IntMean)
colnames(IntMean)<-c("interval", "msteps")
# substitute NA by step mean per interval
for(i in IntMean$interval){
        mydata[is.na(mydata$steps) & mydata$interval==i, 1]<-IntMean[IntMean$interval==i, 2]
}
ManData<-mydata
```

#####Histogram
The histogram illustrates the frequency of steps per day [NA filled with step means per interval].
```{r,  fig.width=5, fig.height=3}
hist(tapply(ManData$steps, ManData$date, sum, na.rm=FALSE),
     main = "Frequency of steps taken per day",
     xlab = "Steps per day",
     ylab = "Frequency",
     breaks = 20,
     border = "blue",
     col="lightblue", 
     xlim=c(0,25000),
     ylim=c(0, 20),
     las=2)
```

#####Mean, Median of the total number of steps taken per day

```{r,  results='asis'}
library(xtable)
StepsByDay<-tapply(ManData$steps , ManData$date, sum, na.rm=TRUE)
stepsMean<-round(mean(StepsByDay), 0)
stepsMedian<-median(StepsByDay)
totSteps_I<-sum(ManData$steps, na.rm = TRUE)
df<-data.frame()
df<-rbind(stepsMean, stepsMedian,totSteps_I)
df<-cbind(df, c(" mean per day ", " median per day ", " total all days "))
colnames(df)<-c(" Values ", " Description ")
xt <- xtable(df)
print(xt, type = 'html', include.rownames = FALSE)
```


By filling in the missing values the total number of steps increased from  570608 to 656704. Mean and median are now very close together. The mean is actually a bit greater that the median. The histrogram looks in its from very similar but increased in the center by about 8000 steps per day.  

##Differences in activity patterns between weekdays and weekends
```{r,  results='asis', fig.width=5,fig.height=3}
library(lattice)
# convert dates into the date-format 
mydata$date <- as.Date(mydata$date ,format="%Y-%m-%d")
# generate a dataframe with weekday names 
mydata$day<-weekdays(mydata$date)
# distinguish between weekday and weekend
daycat<-lapply(mydata$day, function(x) {
        if (x == "Saturday" | x == "Sunday")
                {
                x = "Weekend"}
        else {x = "Weekday"}
}
)
wewd<-data.frame(t(as.data.frame(daycat)))
colnames(wewd)<-"weekday-weekend"
mydata$wd_we<-wewd$`weekday-weekend`

## prep panel plots 
# steps weekend
StepsWeekend<-data.frame(tapply(mydata[mydata$wd_we=="Weekend", ]$steps, mydata[mydata$wd_we=="Weekend", ]$interval, mean))
StepsWeekend<-cbind(data.frame(rownames(StepsWeekend)), StepsWeekend)
colnames(StepsWeekend)<-c("interval", "steps")

#steps weekdays
StepsWeekdays<-data.frame(tapply(mydata[mydata$wd_we=="Weekday", ]$steps, mydata[mydata$wd_we=="Weekday", ]$interval, mean))
StepsWeekdays<-cbind(data.frame(rownames(StepsWeekdays)), StepsWeekdays)
colnames(StepsWeekdays)<-c("interval", "steps")

library(lattice)
df<-data.frame(tapply(mydata$steps, list(mydata$interval, mydata$wd_we), mean))
df$interval <- rownames(df)

wd<-data.frame(df[,c(3,1)])
wd$cat<-"weekday"
colnames(wd)<-c("interval", "steps", "wd_we")
we<-data.frame(df[,c(3,2)])
we$cat<-"weekend"
colnames(we)<-c("interval", "steps", "wd_we")

myw<-rbind(wd,we)
xyplot(as.numeric(myw$steps) ~ as.numeric(myw$interval)| as.factor(myw$wd_we), layout = c(1,2), col = "red", type = "l")
```
